New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions

### 2022-09-08 paper/paper.tex
Went over sections 1, 2, 4, updated notation for the application section.

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications, the more useful methods
for data analysis need to work for arbitrary dimension. A good approah
is to show projections from an arbitrary dimensional space to create
dynamic data visualizations called \emph{tours}. Tours involve views of
high-dimensional (\(p\)) data with low-dimensional (\(d\)) projections.
In his original paper on the grand tour, \citet{As85} provided several
algorithms for tour paths that could theoretically show the viewer the
data \emph{from all sides}. Prior to Asimov's work, there were numerous
preparatory developments including \citet{tukey}'s PRIM-9. PRIM-9 had
user-controlled rotations on coordinate axes, allowing one to manually
tour through low-dimensional projections. (A video illustrating the
capabilities is available through video library of \citet{ASA22}.)
Steering through all possible projections is impossible, unlike Asimov's
tours which allows one to quickly see many, many different projections.
After Asimov there have been many, many tour developments, which are
summarized in \citet{lee2021}.

@@

A manual tour allows the user to alter the coefficients of one (or more)
variables contributing to a \(d-D\) projection. The initial ingredients
are an orthonormal basis (\(A_{p\times d}\)) defining the projection of
the data, and a variable id (\(m \in \{1, ..., p\}\)) specifying which
coefficient will be changed. A method to update the values of the
component (\(m^{th}\) row of \(A_{p\times d}\)) of the controlled
variable \(V_m\) is then needed.

@@

  \begin{enumerate}
  \def\labelenumii{\roman{enumii}.}
  \tightlist
  \item
    Normalise \(\boldsymbol{a}^*_1\) and \(\boldsymbol{a}^*_2\).
  \item
    \(\boldsymbol{a}^*_2 = \boldsymbol{a}^*_2 - <\boldsymbol{a}^*_1\cdot\boldsymbol{a}^*_2>\boldsymbol{a}^*_1\).
  \end{enumerate}
\end{enumerate}

@@

The problem with new simple method is that it is not faithful to the
precise values for \(V_m\) because the orthonormalisation will change
them. There are numerous ways that this can be enforced, and details are
in the Appendix.

@@

In the case of a single orthogonal direction on the projection plane we
can pick a sequence of center points \(\mathbf{c}\) in steps along that
direction to move the slice and fully cover the data space. This no
longer works in higher-dimensional spaces, and we can think of picking
one direction and shifting the slice along the component orthogonal to
the projection plane. In practice we define the center point
\(\mathbf{c}\) explicitly in terms of all \(p\) components and calculate
the slice display following Eq. \ref{eq:slice}.

@@

In addition it is also useful to interactively change the slice
thickness \(h\) (also called the slice radius), in particular to find
the preferred value for exploring the input data. This was implemented
as a slider, and to allow the user to efficiently explore the
interesting range we provide the slider range as an input. For guidance
the estimates of the number of points inside the slice as a function of
the original sample size \(N\) and the number of dimensions \(p\) from
\citet{sectionpursuit} can be used: in case of a uniform distribution
inside a sphere of radius \(R\) a slice with thickness \(h\) will
contain \(N_S\) points, with \begin{equation}
N_S = \frac{N}{2} \left(\frac{h}{R}\right)^{p-2} \left(p - (p-2)\left(\frac{h}{R}\right)^{2}\right).
\label{eq:count}
\end{equation}

@@

We start by first exploring the projections of the model prediction.
Figure \ref{proj1} summarises the process. Through manual rotation of
the view we can get a feeling for where in the space we primarily
predict each of the three species, and we also get a sense of the
difference between the two models. To illustrate this difference we have
manually rotated the projection for the RF model (left plot) to identify
a projection that shows the non-linear but block-type structure that is
typical of this type of model. This particular projection (\(A_1\)) is
exported so that the same projection can be used to show the LDA model
(middle plot) and the actual data (right plot). What can be seen is the
linearity of the LDA model, where the boundaries are linear and oblique
to the variable axes. And, interestingly this particular projection of
the original data shows very distinct clusters of the three species.
That means, the obscuring of the boundaries between groups for both of
the models is driven by what is happening in the orthogonal space to the
plane of the selected projection.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.32\textwidth]{figures/proj1_rf.png}
\includegraphics[width=0.32\textwidth]{figures/proj1_lda.png}
\includegraphics[width=0.32\textwidth]{figures/proj1_data.png}}
\caption{Projection identified using the manual tour, because it reveals an interesting structure in the predictions from the RF model (left). We can clearly see a block structure, while the LDA model (middle) produces linear boundaries. The three groups are nicely separated in this projection of the data (right)}.
\label{proj1}
\end{figure*}

@@

XXXX currently in the notebook \(A_1\) is called A5, and \(A_2\) is
called A6

We continue the investigation by now slicing orthogonally to the
projection \(A_1\). For both models we look at a thin (\(h=0.5\)) slice
through the center, \(S_1^0 = (0,0,0,0)\). At first we explore how
changing the projection, and thus the slice) away from \(A_1\) can help
with understanding the boundary better. For our example, notice that
\(A_1\) does not contain any contribution from the second variable
(\texttt{bd}), so we will first rotate this variable into the view.

Figure \ref{slice1} shows snapshots of the exploration. The rop row is
the initial projection (\(A_1\) and \(S_1^0\)), and the bottom row is a
later projection containing more of \texttt{bd} (call them \(A_2\) and
\(S_2^0\)). The columns correspond to the two models, with Rf on the
left and LDA on the right.

Both models have some overlap at the center, even for this thin slice
(\(S_1^0\)). The second slice (\(S_2^0\)) mostly resolves this overlap
and reveals the primary differences between the models. The boundary
between Adelie (blue) and Chinstrap (yellow) is very similar for both
models. The boundary between Chinstrap and Gentoo (green) is where they
differ.

The RF is almost straight in this view, as something we might expect
from a tree model where splitting occurs on single variables only.
However, it is straight in a combination of the second and third
variable (\texttt{bd} and \texttt{fl}). By examining the scatterplot
matrix, we can understand how this boundary was built: on each of
\texttt{bd} and \texttt{fl} it is possible to cut on a value where most
of the Gentoo penguins are different from the other two speciesfor any
tree, and likely only one is needed. Thus the forest construction is
providing a sample of trees that use one variable or the other variable
to split, producing the blocky boundary on the combination of the two.

@@

@@

We have seen that starting from \(A_1\) using the manual controls to
change the contribution of the second variable we could find a clear
separation boundary indicating the relation between this variable (bill
depth) and the Gentoo penguin species. Instead of rotating to a
different projection, we might also change the view by moving the slice
along one axis in the 4D space. Here we will continue our exploration of
the dependence on \texttt{bd} and move the slice defined by \(A_1\),
\(S_1^c\) to either large positive or negative values
(\(c_{bd} = \pm 1.5\) after centering and scaling). We will label these
slices as \(S_1^{\pm}\). Here we will also look at slices of the
observed data points, using a thicker slice (\(h=1.5\)) to capture
enough points in a given view.

We start by a comparison of the two models and the data distribution in
\(S_1^{+}\), thus the slice is localized towards high values of bill
depth in Figure \ref{slice1p}. We can see that all three slices (the two
models and the data) contain almost no points from the third class
(green, Gentoo), and that the decision boundary between the two models
is very similar.

@@

A more interesting comparison is found for \(S_1^{-}\), thus the slice
localized towards low values of bill depth, shown in Figure
\ref{slice1m}. The RF model (left) predicts all three species within
this slice, with an interesting boundary for the third class (green,
Gentoo). On the other hand the LDA model (middle) predominantly predicts
the third class within the slice, this appears to be enforced through
the linear structure of the model. Looking finally at the thick slice
through the data we see that there are primarily observations from this
class within the slice we can conclude that the two models have filled
in the ``empty'' space (where we do not have any training observations)
in very different ways and according to what we might expect given the
model structure.

@@

Mathematica provides much utility and versatility, such as inbuilt data
visualisation, data manipulation and analysis, dynamic functionality,
and symbolic and numeric computation. Most of this inbuilt functionality
is quite user-friendly and can be found in the Mathematica
documentation; typically, numerous examples are provided within the
documentation, and some possible issues are outlined there. Also, it is
relatively simple to create dynamic objects via the inbuilt commands
Manipulate or DynamicModule. Control objects, such as sliders, locator
panes, and input fields, can then be used on dynamic variables, and when
there is a change in the dynamic variable the dynamic objects which
contain that dynamic variable will be updated. This is the essential
ingredient for this implementation of the manual tour. However, it can
be hard to alter the behaviour of the control objects such as setting
the 1D slider on a different scale (Slider increments add 10 points into
the slice rather than increasing the slice height by 0.1 for example).
\{\textcolor{red}Alex, is this still true with the controls you are
using now?\}

The examples presented in the penguins' notebooks all utilise our
primary new function, SliceDynamic. This function typically accepts
grouped data in the form of a matrix where the second last column
details the name of the group and the last column details the group
index. After specifying the initial slice height and the slice range,
the user is then presented with an interactive display in which the
control objects appear on the left and the slice visulisation appears on
the right. The user can change the orientation of the slice via the
locator pane, which changes the projection matrix; a slider controls the
slice height; and there is an input field, which changes the slice
centre. The user can also change the appearance of the plot by zooming
into the center or changing the point size with sliders provided. The
current projection can be displayed to contrast it with the slice by
ticking a box. The explicit projection matrix can be displayed via
another checkbox. The Slice Tour* notebooks include the additional
functions: ProjectionPlot, Projected2DSliderPlot, and
VisualiseSliceDyanmic. ProjectionPlot is very similar to SliceDynamic,
except it only displays projections and VisualiseSliceDynamic allows you
to discern which points exist inside the slice and outside it. However,
ProjectedLocatorPlot displays the interactive slightly differently.
Several locator panes are displayed above the plot and each one
corresponds to a row in the projection matrix but behaves the same as
ProjectionPlot. \{\textcolor{blue}(this may be easier to implement in
another language (?) )\}

>>>>

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications, the more useful methods
for data analysis need to work for arbitrary dimension. A good approch
is to show projections from an arbitrary dimensional space to create
dynamic data visualizations called \emph{tours}. Tours involve views of
high-dimensional (\(p\)) data with low-dimensional (\(d\)) projections.
In his original paper on the grand tour, \citet{As85} provided several
algorithms for tour paths that could theoretically show the viewer the
data \emph{from all sides}. Prior to Asimov's work, there were numerous
preparatory developments including \citet{tukey}'s PRIM-9. PRIM-9 had
user-controlled rotations on coordinate axes, allowing one to manually
tour through low-dimensional projections. (A video illustrating the
capabilities is available through video library of \citet{ASA22}.)
Steering through all possible projections is impossible, unlike Asimov's
tours which allows one to quickly see many, many different projections.
After Asimov there have been many, many tour developments, which are
summarized in \citet{lee2021}.

@@

A manual tour allows the user to alter the coefficients of one (or more)
variables contributing to a \(d\) dimensional projection. The initial
ingredients are an orthonormal basis (\(A_{p\times d}\)) defining the
projection of the data, and a variable id (\(m \in \{1, ..., p\}\))
specifying which coefficient will be changed. A method to update the
values of the component (\(m^{th}\) row of \(A_{p\times d}\)) of the
controlled variable \(V_m\) is then needed.

@@

  \begin{enumerate}
  \def\labelenumii{\roman{enumii}.}
  \tightlist
  \item
    Normalise \(\boldsymbol{a}^*_1\) and \(\boldsymbol{a}^*_2\).
  \item
    \(\boldsymbol{a}^*_2 = \boldsymbol{a}^*_2 - (\boldsymbol{a}^*_1\cdot\boldsymbol{a}^*_2)\boldsymbol{a}^*_1\).
  \end{enumerate}
\end{enumerate}

@@

The problem with new simple method is that it is not faithful to the
precise values for \(V_m\) because the orthonormalisation will change
them. There are numerous ways that this can be enforced, and details are
in the Appendix.
\textcolor{red}{While they all enforce the same position for the selected variable, they differ in how the remaining variables are adjusted during orthonormalisation.}

@@

A natural starting point is to place \(\mathbf{c}\) in the center of the
data distribution, but shifting it away from the mean can provide
additional insights. In the case of a single orthogonal direction on the
projection plane we can pick a sequence of center points \(\mathbf{c}\)
in steps along that direction to move the slice and fully cover the data
space. This no longer works in higher-dimensional spaces, and we can
think of picking one direction and shifting the slice along the
component orthogonal to the projection plane.

@@

In addition it is also useful to interactively change the slice
thickness \(h\) (also called the slice radius), in particular to find
the preferred value for exploring the input data. For guidance the
estimates of the number of points inside the slice as a function of the
original sample size \(N\) and the number of dimensions \(p\) from
\citet{sectionpursuit} can be used: in case of a uniform distribution
inside a sphere of radius \(R\) a slice with thickness \(h\) will
contain \(N_S\) points, with \begin{equation}
N_S(h, p, R, N) = \frac{N}{2} \left(\frac{h}{R}\right)^{p-2} \left(p - (p-2)\left(\frac{h}{R}\right)^{2}\right).
\label{eq:count}
\end{equation}

In practice we define the center point \(\mathbf{c}\) explicitly in
terms of all \(p\) components and calculate the slice display following
Eq. \ref{eq:slice}.

This was implemented as a slider, and to allow the user to efficiently
explore the interesting range we provide the slider range as an input.

Mathematica provides much utility and versatility, such as inbuilt data
visualisation, data manipulation and analysis, dynamic functionality,
and symbolic and numeric computation. Most of this inbuilt functionality
is quite user-friendly and can be found in the Mathematica
documentation; typically, numerous examples are provided within the
documentation, and some possible issues are outlined there. Also, it is
relatively simple to create dynamic objects via the inbuilt commands
\texttt{Manipulate} or \texttt{DynamicModule}. Control objects, such as
sliders, locator panes, and input fields, can then be used on dynamic
variables, and when there is a change in the dynamic variable the
dynamic objects which contain that dynamic variable will be updated.
This is the essential ingredient for this implementation of the manual
tour. However, it can be hard to alter the behaviour of the control
objects such as setting the 1D slider on a different scale (Slider
increments add 10 points into the slice rather than increasing the slice
height by 0.1 for example). \{\textcolor{red}Alex, is this still true
with the controls you are using now?\}

The examples presented in the penguins' notebooks all utilise our
primary new function, \texttt{SliceDynamic}. This function typically
accepts grouped data in the form of a matrix where the second last
column details the name of the group and the last column details the
group index. After specifying the initial slice height and the slice
range, the user is then presented with an interactive display in which
the control objects appear on the left and the slice visulisation
appears on the right. The user can change the orientation of the slice
via the locator pane, which changes the projection matrix; a slider
controls the slice height; and there is an input field, which changes
the slice centre. The user can also change the appearance of the plot by
zooming into the center or changing the point size with sliders
provided. The current projection can be displayed to contrast it with
the slice by ticking a box. The explicit projection matrix can be
displayed via another checkbox. The Slice Tour* notebooks include the
additional functions: ProjectionPlot, Projected2DSliderPlot, and
VisualiseSliceDyanmic. ProjectionPlot is very similar to SliceDynamic,
except it only displays projections and VisualiseSliceDynamic allows you
to discern which points exist inside the slice and outside it. However,
ProjectedLocatorPlot displays the interactive slightly differently.
Several locator panes are displayed above the plot and each one
corresponds to a row in the projection matrix but behaves the same as
ProjectionPlot. \{\textcolor{blue}(this may be easier to implement in
another language (?) )\}

@@

We start by first exploring the projections of the model prediction.
Figure \ref{proj1} summarises the process. Through manual rotation of
the view we can get a feeling for where in the space we primarily
predict each of the three species, and we also get a sense of the
difference between the two models. To illustrate this difference we have
manually rotated the projection for the RF model (left plot) to identify
a view that shows the non-linear but block-type structure that is
typical of this type of model. This particular projection (\(A_1\)) is
exported so that the same projection can be used to show the LDA model
(middle plot) and the actual data (right plot). What can be seen is the
linearity of the LDA model, where the boundaries are linear and oblique
to the variable axes. And, interestingly this particular projection of
the original data shows very distinct clusters of the three species.
That means, the obscuring of the boundaries between groups for both of
the models is driven by what is happening in the orthogonal space to the
plane of the selected projection.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.32\textwidth]{figures/proj1_rf.png}
\includegraphics[width=0.32\textwidth]{figures/proj1_lda.png}
\includegraphics[width=0.32\textwidth]{figures/proj1_data.png}}
\caption{Projection identified using the manual tour, because it reveals an interesting structure in the predictions from the RF model (left). We can clearly see a block structure, while the LDA model (middle) produces linear boundaries. The three groups are nicely separated in this projection of the data (right).}
\label{proj1}
\end{figure*}

This example can be reproduced with the code in
penguins\_exploring\_manually.nb and a run through is shown in the video
at \url{https://vimeo.com/747585410}.

@@

We continue the investigation by now slicing orthogonally to the
projection \(A_1\). For both models we look at a thin (\(h=0.5\)) slice
through the center, \(\mathbf{c}^0 = (0,0,0,0)\). At first we explore
how changing the projection away from \(A_1\) can help with
understanding the boundary better. For our example, notice that \(A_1\)
does not contain any contribution from the second variable
(\texttt{bd}), so we will first rotate this variable into the view.

Figure \ref{slice1} shows snapshots of the exploration. The top row is
the initial projection (\(A_1\) and \(\mathbf{c}^0\)), and the bottom
row is a later projection containing more of \texttt{bd} (new projection
\(A_2\) while keeping the center fixed at \(\mathbf{c}^0\)). The columns
correspond to the two models, with RF on the left and LDA on the right.

Both models have some overlap at the center for the thin slice based on
\(A_1\) (despite the small value for \(h\)). The second slice (based on
\(A_2\)) mostly resolves this overlap and reveals the primary
differences between the models. The boundary between Adelie (blue) and
Chinstrap (yellow) is very similar for both models. The boundary between
Chinstrap and Gentoo (green) is where they differ.

The RF is almost straight in this view, as something we might expect
from a tree model where splitting occurs on single variables only.
However, it is straight in a combination of the second and third
variable (\texttt{bd} and \texttt{fl}). By examining the scatterplot
matrix, we can understand how this boundary was built: on each of
\texttt{bd} and \texttt{fl} it is possible to cut on a value where most
of the Gentoo penguins are different from the other two species for any
tree, and likely only one is needed. Thus the forest construction is
providing a sample of trees that use one variable or the other variable
to split, producing the blocky boundary on the combination of the two.

@@

This example can be reproduced with the code in
penguins\_slicing\_through\_center.nb and a run through is shown in the
first part of the video at \url{https://vimeo.com/747590472}.

@@

We have seen that starting from \(A_1\) using the manual controls to
change the contribution of the second variable we could find a clear
separation boundary indicating the relation between this variable (bill
depth) and the Gentoo penguin species. Instead of rotating to a
different projection, we might also change the view by moving the slice
along one axis in the 4D space. Here we will continue our exploration of
the dependence on \texttt{bd} and move the slice defined by \(A_1\) to
either large positive or negative values (\(c^{\pm}_{bd} = \pm 1.5\)
after centering and scaling). Thus we define
\(\mathbf{c}^{\pm} = (0, c^{\pm}_{bd}, 0, 0)\). Here we will also look
at slices of the observed data points, using a thicker slice (\(h=1.5\))
to capture enough points in a given view.

We start by a comparison of the two models and the data distribution
shifting to \(\mathbf{c}^{+}\), thus the slice is localized towards high
values of bill depth in Figure \ref{slice1p}. We can see that all three
slices (the two models and the data) contain almost no points from the
third class (green, Gentoo), and that the decision boundary between the
two models is very similar.

@@

A more interesting comparison is found for \(\mathbf{c}^{-}\), thus the
slice localized towards low values of bill depth, shown in Figure
\ref{slice1m}. The RF model (left) predicts all three species within
this slice, with an interesting boundary for the third class (green,
Gentoo). On the other hand the LDA model (middle) predominantly predicts
the third class within the slice, this appears to be enforced through
the linear structure of the model. Looking finally at the thick slice
through the data we see that there are primarily observations from this
class within the slice we can conclude that the two models have filled
in the ``empty'' space (where we do not have any training observations)
in very different ways and according to what we might expect given the
model structure.

@@

This example can be reproduced with the code in
penguins\_shifting\_slice\_center.nb and a run through is shown in the
second part of the video at \url{https://vimeo.com/747590472}.
